Detecting distributional differences between temporal granularities for exploratory time series analysis

### 2021-05-31 paper/main.Rmd
till combination approach

# A distance measure for distributional differences in harmonies {#sec:computation-wpd}

We propose a measure called Weighted Pairwise Distances ($wpd$) to assess structure in the measured variable across two cyclic granularities.

@@

The principle behind the construction of $wpd$ is explained through a simple example explained in Figure \ref{fig:null4by2}. Each of these figures describes a panel with $2$ x-axis categories and $3$ facet levels, but with different designs. Figure \ref{fig:null4by2}a has all categories drawn from N(0, 1) distribution for each facet. It is not an interesting display particularly, as distributions do not vary across x-axis or facet categories. Figure \ref{fig:null4by2}b has x categories drawn from the same distribution, but across facets the distributions are $3$ standard deviations apart. Figure \ref{fig:null4by2}c exhibits an exact opposite situation where distribution between the x-axis categories
are $3$ standard deviations apart, but they do not change across facets. \ref{fig:null4by2}d takes a step further by varying the distribution across both facet and x-axis categories by  $3$ standard deviations. If the panels are to be ranked in order of capturing maximum variation in the measured variable from minimum to maximum, then an obvious choice would be placing (a) followed by (b), (c) and then (d). It might be argued that it is not clear if (b) should precede or succeed (c) in the ranking. Gestalt theory suggests items placed at close proximity can be compared more easily, because people assume that they are in the same group and apart from other groups. With this principle in mind, display (b) is considered less informative as compared to display (c) in emphasizing the distributional differences. The proposed measure $wpd$ is constructed in a way so that it could be used to rank panels of different designs as well as test if a design is interesting. This measure is an estimate of the maximum variation in the measured variable explained by the panel. A higher value of  $wpd$ would indicate that the panel is interesting to look at, whereas a lower value would indicate otherwise.

@@

Consider two cyclic granularities $A$ and $B$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$ with $A$ placed across x-axis and $B$ across facets. Let $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ be a continuous variable observed across $T$ time points. This  data structure with $J$ x-axis levels and $K$ facet levels is referred to as a $(J, K)$ panel. For example, a $(2, 3)$ panel will have cyclic granularities with 2 x-axis levels and 3 facet levels. Let the four elementary designs as described in Figure \ref{fig:null4by2} be $D_{null}$ where there is no difference in distribution of $v$ for $A$ or $B$, $D_{var_f}$ denotes the set of designs where there is difference in distribution of $v$ for $B$ and not for $A$. Similarly, $D_{var_x}$ denotes the set of designs where difference is observed only across A. Finally, $D_{var_{all}}$ denotes those designs for which difference is observed across both $A$ and $B$.

@@

The computation of the distance measure $wpd$ for a panel involves characterizing distributions, computing distances between distributions, choosing a tuning parameter to specify the weightage for different group of distances and summarizing those weighted distances appropriately to estimate maximum variation. Furthermore, the data needs to be appropriately transformed to ensure that the value of $wpd$ detect distributional differences across categories and not across different data generating processes.

@@

Multiple observations of $v$ correspond to the subset  $v_{jk} = \{s: A(s) = j, B(s) = k\}$. The number of observations might vary widely across subsets due to the structure of the calendar, missing observations or uneven locations of events in the time domain. Quantiles of $v_{jk}$'s are chosen as a way to characterize distributions for the category $(a_j, b_k)$ in the paper $\forall j\in \{1, 2, \dots, J\}, k\in \{1, 2, \dots, K\}\}$. The quantile of a distribution with probability $p$ is defined as $Q(p)=F^{-1}(p) = inf\{x: F(x) >p\}$, $0<p< 1$ where $F(x)$ is the distribution function. There are two broad approaches to quantile estimation, viz, parametric and non-parametric. Sample quantiles (@Hyndman1996-ty) are used for estimating population quantiles in a non-parametric setup, which is desirable because of less rigid assumptions made about the nature of the underlying distribution of the data. The default quantile chosen in this paper is percentiles computed for $p = {0.01, 0.02, \dots, 0.99}$, where for example, the $99^{th}$ percentile would be the value corresponding to $p=0.99$ and hence 99% of the observations would lie below that.

@@

One of the most common ways to measure divergence between distributions is the Kullback-Leibler (KL) divergence [@Kullback1951-jy]. The KL divergence denoted by $D(q_1||q_2)$ is a non-symmetric measure of the difference between two probability distributions $q_1$ and $q_2$ and is interpreted as the amount of information lost when $q_2$ is used to approximate $q_1$. Although the KL divergence measures the “distance” between two distributions, it is not a distance measure since it is not symmetric. The Jensen-Shannon divergence [@Menendez1997-in] based on the Kullback-Leibler divergence is symmetric and it always has a finite value. In this paper, the pairwise distances between the distributions of the measured variable are obtained through the square root of the Jensen-Shannon divergence, called Jensen-Shannon distance (JSD) and is defined by,

@@

Furthermore, these distances are distributed as chi-squared with $m$ degrees of freedom (@Menendez1997-in), if the continuous distribution is being discretized with $m$ discrete values. Taking sample percentiles to approximate the integral would mean taking $m = 99$. As the degrees of freedom $m$ get larger, the chi-square distribution approaches the normal distribution.

@@

\noindent Pairwise distances could be within-facets or between-facets. Figure \ref{fig:distance-explain} illustrates how the within-facet or between-facet distances are defined. Pairwise distances are within-facets when $b_{k} = b_{k'}$, that is, between pairs of the form $(a_{j}b_{k}, a_{j'}b_{k})$ as shown in panel (3) of Figure \ref{fig:distance-explain}. If categories are ordered (like all temporal cyclic granularities), then only distances between pairs where $a_{j'} = (a_{j+1})$ are considered (panel (4)). Pairwise distances are between-facets when they are considered between pairs of the form $(a_{j}b_{k}, a_{j}b_{k'})$. Number of between-facet distances would be $^{K}C_2*J$ and number of within-facet distances are $K*(J-1)$ (ordered) and $^{J}C_2*K$ (un-ordered).

@@

A tuning parameter specifying the weightage given to the within-facet or between-facet categories can help to choose between designs like \ref{fig:null4by2}(b) and (c). Following the principle of Gestalt theory, the
relative importance of within-facet and between-facet distances are taken to be $2:1$ and hence $\lambda = \frac{2}{3} = 0.67$. No human experiment is conducted to justify this ratio, however, typically a tuning parameter $\lambda>0.5$ would tend to upweigh the within-facet distances and that with $<0.5$ would upweigh the between-facet distances (refer to the Supplementary section of the paper for more details).

@@

Ideally it is desired that $wpd$ takes a higher value only if there is a significant difference between distributions across facet or x-axis categories, and not because the number of categories $J$ or $K$ is high. Therefore, in order to compare $wpd$ across panels with different levels, the impact of different number of comparisons need to be eliminated through some adjustment. Henceforth, we call the adjusted measure as $wpd_{norm}$. Two approaches for adjusting for the number of comparisons are discussed, both of which are substantiated using simulations. The first one, denoted as $wpd_{perm}$, involves a permutation method to make the distribution of the adjusted distance measure similar for different comparisons. The second one ($wpd_{glm-scaled}$) fits a model to represent the relationship between $wpd_{raw}$ and number of comparisons and defines the adjusted measure as the residual from the model.

- _Permutation approach:_ This method is somewhat similar in spirit to bootstrap or permutation tests, where the goal is to test the hypothesis that the groups under study have identical distributions. This method, in essence, accomplishes a different goal of making the distribution of the different groups (panels in our case) same under $D_{null}$. The values of $wpd_{raw}$ is computed on many permuted ($nperm$) data sets for the $l^{th}$ panel and stored in $wpd_{perm-data}$. $nperm=100$ is considered and the adjusted distance measure
$wpd_{perm}$ is computed as follows:

$$wpd_{perm} = \frac{(wpd_{raw} - \bar{wpd_{perm-data}})}{sd(wpd_{perm-data})}$$
where $\bar{wpd_{perm-data}}$ and $sd(wpd_{perm-data})$ are the mean and standard deviation of $wpd_{perm-data}$ respectively. Standardizing  $wpd$ in the permutation approach ensures that the distribution of $wpd_{perm}$ has the same $mean = 0$ and $sd = 1$ across all comparisons under $D_{null}$. While this works successfully to make the location and scale similar across different $nx$ and $nfacet$ (as seen in Figure \ref{fig:norm}), it is computationally heavy and time consuming, and hence less user friendly when being actually used in practice. Hence, another approach to adjustment, with potentially less computational time, is proposed.

- _Modeling approach:_  $wpd_{raw}$ being a Jensen-Shannon distance, follows a Chi-square distribution (@JSD), which is a special case of Gamma distribution. Hence, a Gamma generalized linear model (GLM) for $wpd_{raw}$ is fitted with number of comparisons as the explanatory variable, which is of the following form:

- _Combination approach:_ The simulation results show that adjustment through the modeling approach leads to very similar distribution across high $nx$ and $nfacet$ (higher than $5$) and not so much for lower $nx$ and $nfacet$. Hence, the computational load of permutation approach could be alleviated by using the modeling approach for the higher $nx$ and $nfacet$, however, it is important that we use the permutation approach for lower $nx$ and $nfacet$. The adjusted variables from the two approaches have to be brought to the same scale so that for smaller categories, permutation approach is used and for larger categories, modeling approach is used. The adjusted measure from the combination approach, denoted by $wpd_{norm}$ is defined as follows:

>>>>

# Proposed distance measure {#sec:computation-wpd}

We propose a measure called Weighted Pairwise Distances ($wpd$) to detect distributional differences in the measured variable across cyclic granularities.

@@

The principle behind the construction of $wpd$ is explained through a simple example explained in Figure \ref{fig:null4by2}. Each of these figures describes a panel with $2$ x-axis categories and $3$ facet levels, but with different designs. Figure \ref{fig:null4by2}a has all categories drawn from N(0, 1) distribution for each facet. It is not an interesting display particularly, as distributions do not vary across x-axis or facet categories. Figure \ref{fig:null4by2}b has x categories drawn from the same distribution, but across facets the distributions are $3$ standard deviations apart. Figure \ref{fig:null4by2}c exhibits an exact opposite situation where distribution between the x-axis categories
are $3$ standard deviations apart, but they do not change across facets. Figure \ref{fig:null4by2}d takes a step further by varying the distribution across both facet and x-axis categories by  $3$ standard deviations. If the panels are to be ranked in order of capturing maximum variation in the measured variable from minimum to maximum, then an obvious choice would be placing (a) followed by (b), (c) and then (d). It might be argued that it is not clear if (b) should precede or succeed (c) in the ranking. Gestalt theory suggests items placed at close proximity can be compared more easily, because people assume that they are in the same group and apart from other groups. With this principle in mind, display (b) is considered less informative as compared to display (c) in emphasizing the distributional differences. Considering one cyclic granularity, we would have only two design choices similar to (a) and (c), corresponding to no difference and significant differences between categories of that cyclic granularity only. The proposed measure $wpd$ is constructed in a way so that it could be used to rank panels of different designs as well as test if a design is interesting. This measure is aimed to be an estimate of the maximum variation in the measured variable explained by the panel. A higher value of  $wpd$ would indicate that the panel is interesting to look at, whereas a lower value would indicate otherwise.

@@

\noindent Let the number of cyclic granularities considered in the display be $m$. The notations and methodology are described in detail for $m=2$. But it can be easily extended to $m>2$. Consider two cyclic granularities $A$ and $B$, such that $A = \{ a_j: j = 1, 2, \dots, J\}$ and $B = \{ b_k: k = 1, 2, \dots, K\}$ with $A$ placed across x-axis and $B$ across facets. Let $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ be a continuous variable observed across $T$ time points. This  data structure with $J$ x-axis levels and $K$ facet levels is referred to as a $(J, K)$ panel. For example, a $(2, 3)$ panel will have cyclic granularities with 2 x-axis levels and 3 facet levels. Let the four elementary designs as described in Figure \ref{fig:null4by2} be $D_{null}$ where there is no difference in distribution of $v$ for $A$ or $B$, $D_{var_f}$ denotes the set of designs where there is difference in distribution of $v$ for $B$ and not for $A$. Similarly, $D_{var_x}$ denotes the set of designs where difference is observed only across A. Finally, $D_{var_{all}}$ denotes those designs for which difference is observed across both $A$ and $B$. $m = 1$ is a special case of $m = 2$ with $J = 1$.

@@

The computation of the distance measure $wpd$ for a panel involves characterizing distributions, computing distances between distributions, choosing a tuning parameter to specify the weightage for different group of distances and summarizing those weighted distances appropriately to estimate maximum variation. Furthermore, the data needs to be appropriately transformed to ensure that the value of $wpd$ emphasizes detection of distributional differences across categories and not across different data generating processes.

@@

Multiple observations of $v$ correspond to the subset  $v_{jk} = \{s: A(s) = j, B(s) = k\}$. The number of observations might vary widely across subsets due to the structure of the calendar, missing observations or uneven locations of events in the time domain. in this paper, quantiles of $v_{jk}$'s are chosen as a way to characterize distributions for the category $(a_j, b_k)$, $\forall j\in \{1, 2, \dots, J\}, k\in \{1, 2, \dots, K\}\}$. The quantile of a distribution with probability $p$ is defined as $Q(p)=F^{-1}(p) = inf\{x: F(x) >p\}$, $0<p< 1$ where $F(x)$ is the distribution function. There are two broad approaches to quantile estimation, viz, parametric and non-parametric. Sample quantiles (@Hyndman1996-ty) are used for estimating population quantiles in a non-parametric setup, which is desirable because of less rigid assumptions made about the nature of the underlying distribution of the data. The default quantile chosen in this paper is percentiles computed for $p = {0.01, 0.02, \dots, 0.99}$, where for example, the $99^{th}$ percentile would be the value corresponding to $p=0.99$ and hence 99% of the observations would lie below that.

@@

One of the most common ways to measure divergence between distributions is the Kullback-Leibler (KL) divergence [@Kullback1951-jy]. The KL divergence denoted by $D(q_1||q_2)$ is a non-symmetric measure of the difference between two probability distributions $q_1$ and $q_2$ and is interpreted as the amount of information lost when $q_2$ is used to approximate $q_1$. The KL divergence, however, is not symmetric and hence can not be considered as a "distance" measure. The Jensen-Shannon divergence [@Menendez1997-in] based on the Kullback-Leibler divergence is symmetric and has a finite value. Hence, in this paper, the pairwise distances between the distributions of the measured variable are obtained through the square root of the Jensen-Shannon divergence, called Jensen-Shannon distance (JSD) and is defined by,

@@

<!-- Furthermore, these distances are distributed as chi-squared with $m$ degrees of freedom (@Menendez1997-in), if the continuous distribution is being discretized with $m$ discrete values. Taking sample percentiles to approximate the integral would mean taking $m = 99$. As the degrees of freedom $m$ get larger, the chi-square distribution approaches the normal distribution. -->

@@

\noindent Pairwise distances could be within-facets or between-facets. Figure \ref{fig:distance-explain} illustrates how they are defined. Pairwise distances are within-facets when $b_{k} = b_{k'}$, that is, between pairs of the form $(a_{j}b_{k}, a_{j'}b_{k})$ as shown in panel (3) of Figure \ref{fig:distance-explain}. If categories are ordered (like all temporal cyclic granularities), then only distances between pairs where $a_{j'} = (a_{j+1})$ are considered (panel (4)). Pairwise distances are between-facets when they are considered between pairs of the form $(a_{j}b_{k}, a_{j}b_{k'})$. Number of between-facet distances would be $^{K}C_2*J$ and number of within-facet distances are $K*(J-1)$ (ordered) and $^{J}C_2*K$ (un-ordered).

@@

A tuning parameter specifying the weightage given to the within-facet or between-facet categories can help to choose between designs like \ref{fig:null4by2}(b) and (c). The tuning parameters should be chosen such that $\sum_{i=1}^m \lambda_i = 1$. When $m=2$, following the principle of Gestalt theory, $\lambda = \frac{2}{3} = 0.67$ is chosen to put a relative weightage of $2:1$ for within-facet and between-facet distances. No human experiment is conducted to justify this ratio, however, typically a tuning parameter $\lambda>0.5$ would tend to upweigh the within-facet distances and that with $<0.5$ would upweigh the between-facet distances (refer to the Supplementary section of the paper for more details).

@@

Ideally it is desired that $wpd$ takes a higher value only if there is a significant difference between distributions across categories, and not because the number of categories $J$ or $K$ is high. That is, under designs like $D_{null}$, the distribution of the $wpd$ values should not differ for 
different number of categories. Only then the distance measure could be compared across panels with different levels. This calls for an adjusted measure, which normalizes for the different number of comparisons. We denote it by $wpd_{norm}$. Two approaches for adjusting the number of comparisons are discussed, both of which are substantiated using simulations. The first one defines an adjusted measure $wpd_{perm}$ based on permutation method to remove the effect of different comparisons. The second approach fits a model to represent the relationship between $wpd_{raw}$ and number of comparisons and defines the adjusted measure ($wpd_{glm-scaled}$) as the residual from the model.

_Permutation approach_

This method is somewhat similar in spirit to bootstrap or permutation tests, where the goal is to test the hypothesis that the groups under study have identical distributions. This method accomplishes a different goal of finding the null distribution for different groups (panels in our case) and standardizing the raw values using that distribution. The values of $wpd_{raw}$ is computed on many ($nperm$) permuted data sets stored in $wpd_{perm-data}$. Then $wpd_{perm}$ is computed as follows:

\begin{align*}
wpd_{perm} = \frac{(wpd_{raw} - \bar{wpd_{perm-data}})}{sd(wpd_{perm-data})}
\end{align*}

\noindent where $\bar{wpd_{perm-data}}$ and $sd(wpd_{perm-data})$ are the mean and standard deviation of $wpd_{perm-data}$ respectively. Standardizing  $wpd$ in the permutation approach ensures that the distribution of $wpd_{perm}$ has the same $mean = 0$ and $sd = 1$ across all comparisons under $D_{null}$. While this works successfully to make the location and scale similar across different $nx$ and $nfacet$, it is computationally heavy and time consuming, and hence less user friendly when being actually used in practice. Hence, another approach to adjustment, with potentially less computational time, is proposed.

<!-- for the $l^{th}$ panel and -->

_Modeling approach_

In this approach, a Gamma generalized linear model (GLM) for $wpd_{raw}$ is fitted with number of comparisons as the explanatory variable. Since, $wpd_{raw}$ is a Jensen-Shannon distance, it follows a Chi-square distribution (@Menendez1997-in), which is a special case of Gamma distribution. Furthermore, the mean response is bounded, since any JSD is bounded by $1$ given that base $2$ logarithm is used (@Lin1991-pt). Hence, by @Faraway2016-uk, an inverse link is used for the model, which is of the following form:

$$y = a+b*log(z) + e$$ ,

\noindent where $y = wpd_{raw}$, $z = (nx*nfacet)$ is the number of groups and $e$ are idiosyncratic errors. Let $E(y) = \mu$ and $a + b*log(z) = g(\mu)$ where $g$ is the link function. Then $g(\mu)=  1/\mu$ and $\hat \mu  = 1/(\hat a + \hat b log(z))$. The residuals from this model $(y-\hat y) = (y-1/(\hat a + \hat b log(z)))$ would be expected to have no dependency on $z$. Thus, $wpd_{glm}$ is chosen as the residuals from this model and is defined as:
$wpd_{glm} = wpd_{raw} - 1/(\hat a + \hat b*log(nx*nfacet))$.

_Combination approach_

The simulation results (\ref{Sec:}) show that adjustment through the modeling approach leads to very similar distribution across high $nx$ and $nfacet$ (higher than $5$) and not so much for lower $nx$ and $nfacet$. Hence, the computational load of permutation approach is alleviated by using the modeling approach for the higher $nx$ and $nfacet$, however, it is important that we use the permutation approach for lower $nx$ and $nfacet$. The adjusted variables from the two approaches have to be brought to the same scale so that for smaller categories, permutation approach is used and for larger categories, modeling approach is used. The adjusted measure from the combination approach, denoted by $wpd_{norm}$ is defined as follows:
